---
title: "PA1_template"
output: html_document
---


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(sqldf)
library(graphics)
library(ggplot2)
library(timeDate)
setwd("~/Documents/Coursera - Data Science/Reproducible Research/Wk 1")
rawdata <- read.csv("activity.csv", header=TRUE, sep=",")
#remove missing values
clean_data <- na.omit(rawdata)
```

###What is mean total number of steps taken per day?

1. following code calculates the total number of steps taken per day

```{r, results='hide', message=FALSE}
#add the number of steps for each day and aggregate the data by each day
step_sum <-sqldf("select date, sum(steps) as sumsteps from clean_data group by date")
```

2. Following code snippet creates the histogram of the total number of steps taken each day

```{r, results='asis'}
#create the histogram
hist_graph <- hist(step_sum$sumsteps, main="Histogram for total steps in a day",xlab="Steps per day",ylab="Frequency")
```

3. Following code snippet calculates the mean and median of total number of steps taken per day
```{r, results='hide'}
#calculate mean and median for daily total number of steps
daily_mean <- as.integer(mean(step_sum$sumsteps))
daily_median <- as.integer(median(step_sum$sumsteps))
```

Daily mean is `r daily_mean`
Daily median is `r daily_median`


###What is the average daily activity pattern?

1. Following code snippet creates the time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r}
avg_daily_act <- sqldf("select interval, avg(steps) as avg_steps from clean_data group by interval")
max_avg_daily_act <- sqldf("select interval, max(avg_steps) as max_avg from avg_daily_act")
max_interval <- max_avg_daily_act$interval
max_avg_steps <- as.integer(max_avg_daily_act$max_avg)
ggplot(avg_daily_act, aes(interval, avg_steps))  + geom_line() 
```

2. Interval `r max_interval` has the maximum number of steps `r max_avg_steps` on average across all days in the dataset


###Imputing missing values

```{r, echo=TRUE}
missingvalues <- sum(is.na(rawdata$steps))
```

There are `r missingvalues` rows with missing values in the dataset

Replace the missing values in the data set with the mean of all the steps
```{r,echo=TRUE}
clean_data_1 <- rawdata
clean_data_1$steps[which(is.na(clean_data_1$steps))] <- mean(clean_data$steps)
```

Get the total number of steps per day and create a histogram
```{r,echo=TRUE}
step_sum <-sqldf("select date, sum(steps) as sumsteps from clean_data_1 group by date")
#png("plot1.png",height=600,width = 800)
par(mar = rep(2, 4))
hist_graph<-hist(step_sum$sumsteps, main="Histogram for total steps in a day",xlab="Steps per day",ylab="Frequency")
```

Calculate the new mean and median of the total number of steps per day
```{r,echo=TRUE}
daily_mean_1 <- as.integer(mean(step_sum$sumsteps))
daily_median_1 <- as.integer(median(step_sum$sumsteps))
```
New mean is `r daily_mean_1` and new median is `r daily_median_1`


Create a vector that stores if the date in the cleaned date is a weekday or a weekend, then append that vector to the cleaned data set

```{r,echo=TRUE}
wkdy <- vector(mode="logical",length=nrow(clean_data_1))
rowcount <- nrow(clean_data_1)
rownumber <- 0

for (rownumber in 1:rowcount) {
 if (isWeekday(clean_data_1[rownumber,2])) {
   wkdy[rownumber] <- "weekday"
  }
  if (isWeekend(clean_data_1[rownumber,2])){
    wkdy[rownumber] <- "weekend"
   }
}

clean_data_1 <- cbind(clean_data_1,wkdy)
```

Calculate the averages across intervals on weekdays and weekends
```{r,echo=TRUE}
avg_weekday <- sqldf("select interval, avg(steps) as avg_steps, wkdy from clean_data_1 where wkdy='weekday' group by interval")
avg_weekend <- sqldf("select interval, avg(steps) as avg_steps, wkdy from clean_data_1 where wkdy='weekend' group by interval")
```

Plot of avg across intervals on weekdays
```{r,echo=TRUE}
ggplot(avg_weekday, aes(interval, avg_steps))  + geom_line()
```


Plot of avg across intervals on weekends
```{r,echo=TRUE}
ggplot(avg_weekend, aes(interval, avg_steps))  + geom_line()
```

